Handling String-Number Conversion in String Constraint Solving

### 2019-11-18 draft.tex
Update draft.tex

Symbolic execution is very popular technique that allows the programmers to check the feasibility of a path in a given program (i.e., determining the value of the inputs under which the given path can be executed) \cite{????}).
The path feasibility problem is usually solved by a reduction to the satisfiability problem of a formula. More precisely, program statements in the path are usually translated to equivalent constraints in static single assignment (SSA) form and then solved by \emph{satisfiability module theory (SMT)}  solvers. The types of constraints that needed to be solved by the SMT solvers depend on the type of program variables. Therefore, there is a need for SMT solvers that support different combinations of theories (i.e., depending on the types of the manipulated variables).

Among all types of program variables, \emph{string data type}  is omnipresent in modern programming languages.  Various infamous security vulnerabilities such as injection and cross-site scripting attack are caused by malicious string values.  Therefore, string constraint solving has received considerable attention in the constraint solving community.
Operations such as \emph{equality constraints} (e.g. $x.y = y.x$), \emph{regular constraints} (e.g., $x \in (a.b)^*$), and \emph{integer constraints} (e.g., $|x|-|y|>3$), are widely supported by most state-of-the-art string constraint solvers such as Z3  \cite{?}, CVC4  \cite{?}, Z3Str3  \cite{?}, and Trau+ \cite{?}. 
 However,   solvers have limited support for operations that convert between strings and integers; string length operations are sometimes well supported, but not parsing a string as an integer or turning an integer value into its string form.  When there is support, it is often very limited. Thus, $n = \sti{\tt{"03"}}$  or $x = \its{5}$  are handled but $n = \sti{x}$ (i.e., converting the string $x$ to the integer $n$) and  $x = \its{n}$ (i.e.,  converting the integer $n$ to the string $x$)
are ones of the core operations that require stronger solver support.

@@

Currently, even for a simple constraint like the above one, most solvers already fail to provide a correct answer. Therefore, there is an urgent need to efficient techniques for solving the combination of above constraints. To make matters worse, many scripting languages, of which JavaScript is the most notable ---it powers most interactive content on the Web, and increasingly server-side code with Node.js---, has conversions between strings and integers embedded in the core of its semantics. To see how, consider the following program:

@@

A casual glance at the above code reveal no use of strings at all, but the semantics of field access is somewhat unusual in JavaScript: the arrays are indexed by strings, and numeric indexes are converted to strings. This conversion is mandated explicitly by JavaScript semantics: the 2019 edition of ECMAScript requires that $ToPropertyKey$ be called on the element expression (\S{12.3.2.1}), and $ToPropertyKey$ calls {\tt{ToString}} on that value in all but special cases (\S{7.1.14}). Any faithful symbolic execution of JavaScript must handle such conversions for even basic array operations to work correctly; consider the following code snippet that manipulates an array \texttt{x}, with its value shown on the right:

@@

Here \texttt{x[3]} in line 2,  \texttt{x[03]} in line 3, and \texttt{x["3"]} in line 4 all denote the same array element of \texttt{x["3"]}, but \texttt{x["03"]} denotes a completely different element. So na\"ive idea of modeling array index with integers will not work -- it cannot distinguish the indexes \texttt{"3"} and \texttt{"03"}.

If array indexes are modeled as strings, we need to face the problem of handling of arithmetics in the array index. Let us look at the case of line 6, we need to update the value of \texttt{x["03"-1]}. The evaluation of the expression \texttt{"03"-1} involves an implicit type conversion from the string \texttt{"03"} to an integer value $3$ due to the \texttt{-} (minus) operation. The result of the evaluation of \texttt{"03"-1} is the integer $2$, which is then converted back to string \texttt{"2"} and used as the array index. Hence \texttt{x["03"-1]} means the array element of \texttt{x["2"]}. Even for a simple example like this, the conversion between string and number is unavoidable. This is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code.

@@

The Line 4 of the code updates the length of \texttt{x} to $13$ and hence the \texttt{resetArray(x)} will update \texttt{x} to an array with $13$ zeros.
Now suppose that the Line 4 of the code above is replace to {\tt{x[k] = 5}}, where $\mathtt{k}$ is an user input string and we want to find an input such that $\mathtt{x[15]=0}$. In this case, one would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. More precisely, in the corresponding string constraint encoding, we use a shadow variable $xLen$ to remember the length of $x$ and update its value after all array assignment. For an assignment $\mathtt{x[k] = v}$, we update the array value and length using the following string constraint (combined the $select$ function from SMT array theory): $x_{i+1} = store(x_i,k, v)$ and $xLen_{i+1} = ite( xLen_{i}> \sti{k}, xLen_{i}, \sti{k})$.
This is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code.

@@

Solving string constraint with string-number conversion is a very challenging problem. From the theoretical point of view, this problem is already proven to be undecidable~\cite{day2018satisfiability}. In practice, in our experiments, the state-of-the-art string constraint solvers provide only limited support to string-number conversion.

In this paper, we adopt a framework that handles string constraints with string-number conversion efficiently. Due to the high complexity of solving the problem exactly, our framework uses both over and under-approximation of the string constraints. The over-approximation is for proving UNSAT and the under-approximation is for proving SAT of the string constraint. Both over and under-approximation should fail in a decidable fragment of string constraints that can be efficiently solved.

>>>>

Symbolic execution is a very popular technique that allows the programmers to check the feasibility of a path in a given program (i.e., determining the value of the inputs under which the given path can be executed) \cite{????}).
The path feasibility problem is usually solved by a reduction to the satisfiability problem of a formula. More precisely, program statements in the path are usually translated to equivalent constraints in static single assignment (SSA) form and then solved by \emph{satisfiability module theory (SMT)}  solvers. The types of constraints that needed to be solved by the SMT solvers depend on the type of program variables. Therefore, there is a need for SMT solvers that support different combinations of theories (i.e., depending on the types of the manipulated variables).

Among all types of program variables, \emph{string data type}  is omnipresent in modern programming languages.  Various infamous security vulnerabilities such as injection and cross-site scripting attack are caused by malicious string values.  Therefore, string constraint solving has received considerable attention in the constraint solving community.
Operations such as \emph{equality constraints} (e.g. $x.y = y.x$), \emph{regular constraints} (e.g., $x \in (a.b)^*$), and \emph{integer constraints} (e.g., $|x|-|y|>3$), are widely supported by most state-of-the-art string constraint solvers such as Z3  \cite{?}, CVC4  \cite{?}, Z3Str3  \cite{?}, and Trau+ \cite{?}. 
 However,   solvers have limited support for operations that convert between strings and integers; string length operations are sometimes well supported, but not parsing a string as an integer or turning an integer value into its string form.  When there is support, it is often very limited. Thus, $n = \sti{\\texttt{"2"}}$  or $x = \its{5}$  are handled but $n = \sti{x}$ (i.e., converting the string $x$ to the integer $n$) and  $x = \its{n}$ (i.e.,  converting the integer $n$ to the string $x$)
are ones of the core operations that require stronger solver support.

@@

Currently, even for a simple constraint like the above one, most solvers already fail to provide a correct answer. Therefore, there is an urgent need for developing efficient techniques for solving the combination of above constraints. To make matters worse, many scripting languages, of which JavaScript is the most notable ---it powers most interactive content on the Web, and increasingly server-side code with Node.js---, has conversions between strings and integers embedded in the core of its semantics. To see how, consider the following program:

@@

A casual glance at the above code reveal no use of strings at all, but the semantics of field access is somewhat unusual in JavaScript: the arrays are indexed by strings, and numeric indexes are converted to strings. This conversion is mandated explicitly by JavaScript semantics: the 2019 edition of ECMAScript \ref{?}requires that $ToPropertyKey$ be called on the element expression (\S{12.3.2.1}), and $ToPropertyKey$ calls {\tt{ToString}} on that value in all but special cases (\S{7.1.14}). Any faithful symbolic execution of JavaScript must handle such conversions for even basic array operations to work correctly; consider the following code snippet that manipulates an array \texttt{x}, with its value shown on the right:

@@

Here \texttt{x[3]} in line 2,  \texttt{x[03]} in line 3, and \texttt{x["3"]} in line 4 all denote the same array element of \texttt{x["3"]}, but \texttt{x["03"]} denotes a completely different element (which is stored at the index \texttt{"03"} of the array). So na\"ive idea of modeling array index with integers will not work -- it cannot distinguish the indexes \texttt{"3"} and \texttt{"03"}.

If array indexes are modelled as strings, we need to face the problem of handling of arithmetics in the array index. Let us look at the case of line 6, we need to update the value of \texttt{x["03"-1]}. The evaluation of the expression \texttt{"03"-1} involves an implicit type conversion from the string \texttt{"03"} to an integer value $3$ due to the \texttt{-} (minus) operation. The result of the evaluation of \texttt{"03"-1} is the integer $2$, which is then converted back to string \texttt{"2"} and used as the array index. Hence \texttt{x["03"-1]} means the array element of \texttt{x["2"]}. Even for a simple example like this, the conversion between string and number is unavoidable. This is a rather basic array operation in Javascript, and so having only limited capability for handling string-number conversion operations  will cripple any analysis of non-trivial JavaScript code.

@@

The Line 4 of the code updates the length of \texttt{x} to $13$ and hence the \texttt{resetArray(x)} will update \texttt{x} to an array with $13$ zeros.
Now suppose that the Line 4 of the code above is replace to {\tt{x[k] = 5}}, where $\mathtt{k}$ is an user input string and we want to find an input such that $\mathtt{x[15]=0}$. In this case, one would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. More precisely, in the corresponding string constraint encoding, we use a shadow variable $xLen$ to remember the length of $x$ and update its value after all array assignment. For an assignment $\mathtt{x[k] = v}$, we update the array value and length using the following string constraint (combined the $select$ function from SMT array theory): $x_{i+1} = store(x_i,k, v)$ and $xLen_{i+1} = ite( xLen_{i}> \sti{k}, xLen_{i}, \sti{k})$.
This is a rather straightforward array operation, and so  having only limited capability for handling str/num conversion. will cripple any model of non-trivial JavaScript code.

@@

Solving string constraint with string-number conversion is a very challenging problem. From the theoretical point of view, this problem is already proven to be undecidable~\cite{day2018satisfiability}. In practice, in our experiments (Section \ref{section:evaluation}), the state-of-the-art string constraint solvers provide only limited support to string-number conversion.

In this paper, we propose a framework that handles string constraints with string-number conversion efficiently. Due to the high complexity of solving the exact problem, our framework uses both over and under-approximation of the string constraints. The over-approximation is for proving UNSAT and the under-approximation is for proving SAT of the string constraint. Both over and under-approximation  fall in a decidable fragment of string constraints that can we can efficiently solved.
